\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-02-02 slides/IASD/pres.tex
Update the graphical representation
Update the likelihood
Update the predictive marker and minor refactore the slide

\begin{frame}{Submodels}
\begin{itemize}
\item Graphical representation of a joint model of a time-to-event submodel and K-multivariate longitudinal outcomes submodel
\includegraphics[scale=0.42]{figures/lights_graphical_representation.png}
\end{itemize}

\end{frame}
\subsection{Notations}

@@

\footnotesize
\begin{itemize}
  \item \scriptsize $\theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big) \in \R^\vartheta$
  \item $f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\}$ with $\Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}$
  \item Survival part:
  \begin{align*}f(t_i, \delta_i| b_i, G_i = k ; \theta) = &\big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}
  \end{align*}
  \item Then, the likelihood writes
  \begin{align*}\ell_n(\theta) = n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} &\pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) \\ & \times f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i
  \end{align*}
\end{itemize}

@@

\begin{itemize}
  \item Penalized objective
  \begin{equation*}
  \ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
  \end{equation*}
  with the elasticnet penalty \[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \] and the sparse group lasso penalty \[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \]
  \item Resulting optimization problem \[\hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta)\]
\end{itemize}

@@

\subsection{QNMCEM}

\begin{frame}{QNMCEM algorithm (1/2)}

@@

\begin{itemize}
  \item $\ell_n^\text{comp}(\theta) = \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}})$
\end{itemize}

\begin{block}{Monte Carlo E-step}
\begin{itemize}
  \item $\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$
  \item Requires to compute expectations of the form
  \footnotesize
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \] 
\small for different functions $g$, where we denote 
\[\pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] \]
  \item Monte Carlo approximations used for untractable integrals
\end{itemize}
\end{block}

@@

\begin{frame}{QNMCEM algorithm (2/2)}

@@

\begin{block}<1->{Quasi-Newton M-step}
\begin{itemize}
  \item \tiny $\theta^{(w+1)} \in \argmin_{\theta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}$
  \item $D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i]$
  \item \tiny $R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]$
  \item $\beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k)$
  \item $P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$
  \item $\xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}$
  \item L-BFGS-B to solve the problem
  \item Proximal gradient method to estimate $\gamma_k^{(w+1)}$
  \item Predictive marker \scriptsize $\hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}$, which is an estimate of $\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i]$
\end{itemize}
\end{block}

>>>>

\begin{frame}{Submodels}
\begin{itemize}
\item Graphical representation of a joint model of a time-to-event submodel and K-multivariate longitudinal outcomes submodel
\includegraphics[scale=0.29]{figures/lights_graphical_representation.png}
\end{itemize}
\end{frame}
\subsection{Notations}

@@

\footnotesize
\begin{itemize}
  \item<1-> \scriptsize $\theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big) \in \R^\vartheta$
  \item<2-> $f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\}$ with $\Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}$
  \item<3-> Survival part:
  \begin{align*}f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) = &\big[\lambda(t_i| G_i = k, \Psi_i)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s| G_i = k, \Psi_i) \dd s \Big\}
  \end{align*}
  \item<4-> Then, the likelihood writes
  \begin{align*}
  \ell_n(\theta) &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \int_{\R^r} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k, \Psi_i ; \theta) \\ & \qquad\qquad\qquad\qquad \times f(y_i | b_i, G_i = k, \Psi_i; \theta) f(b_i ; \theta) \dd b_i \\
  &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | G_i = k; \theta) \dd b_i
  \end{align*}
\end{itemize}

@@

\begin{itemize}
  \item<1-> Penalized objective
  \begin{equation*}
  \ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
  \end{equation*}
  with the elasticnet penalty \[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \] and the sparse group lasso penalty \[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \]
  \item<2-> Resulting optimization problem \[\hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta)\]
\end{itemize}

@@

\subsection{QNEM}

\begin{frame}{QNEM algorithm (1/2)}

@@

\begin{itemize}
  \item<1-> $\ell_n^\text{comp}(\theta) = \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}})$
\end{itemize}

\begin{block}{E-step}
\begin{itemize}
  \item<2-> $\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$
  \item<3-> Requires to compute expectations of the form
  \footnotesize
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \] 
\small for different functions $g$, where we denote 
\[\pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] \]
\end{itemize}
\end{block}

@@

\begin{frame}{QNEM algorithm (2/2)}

@@

\begin{block}<1->{Quasi-Newton M-step}
\begin{itemize}
  \item<1-> \tiny $\theta^{(w+1)} \in \argmin_{\theta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}$
  \item<2-> $D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i]$
  \item<3-> \tiny $R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]$
  \item<4-> $\beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k)$
  \item<5-> $P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$
  \item<6-> $\xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}$
  \item<7-> L-BFGS-B to solve the problem
  \item<8-> Proximal gradient method to estimate $\gamma_k^{(w+1)}$
  \item<9-> Predictive marker \scriptsize $\hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k, \Psi_i ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k, \Psi_i ; \hat \theta)}$, which is an estimate of $\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i]$
\end{itemize}
\end{block}
